Controlling the rheological properties of well-defined polyelectrolytes hydrogels by fine tuning their ion pairing process

### 2016-01-04 PALSE/ion_pairing_paper.tex
bibliography (generated by Mendeley)

\section{Introduction}
Since the pioneering work of Wichterle et al.,1 which established their relevance in a biomedical context, polymer based hydrogels have never ceased to be a very active field of research.2, 3 They have recently experienced a burst of interest among the biomedical community as controlled drug delivery cargos or scaffolds for wound healing or tissues repair.4 Polyelectrolytes are being extensively used in this particular context.5 Their gel formation mechanism often involves reversible-by-nature electrostatic interactions, which can be used for instance to trigger ``smart'' release of bioactive substances.6 Besides, poly(cationic) gel have been reported to combine scaffold properties for cell’s adherence and growth with antimicrobial activity, and are therefore increasingly used in clinical applications.3, 7 In this framework, injectability is a key feature, as it provides an easier way to gel delivery in vivo; thus, reversible shear thinning biocompatible hydrogels are a particularly sought after class of materials.8
We have recently reported on a highly reversible poly(cation) based hydrogel,9 which formation relies on an new concept of supramolecular  electrostatic interaction. Briefly, the atom transfer radical polymerisation (ATRP) is initiated by a phosphonate terminated ATRP initiator, eventually affording an anion terminated poly(cationic) polymer. The resulting poly(cationic) polymer, when dispersed in an aqueous medium, stabilizes hydrogel formation through the occurrence of a head(anion)-to-body(cations) supramolecular network. This highly dynamic electrostatic network provides the resulting gel with spectacular mechanical and self-healing properties. Moreover the interaction can be disrupted by various chemical stimuli, such as pH or ionic strength.
In the present report, we show that the mechanical/rheological properties of such smart hydrogels can be tuned to an extremely large extent by playing with the nature (hardness/softness) of the cationic repeating unit and the ion-paring strength of the associated counterion (\ce{F-}/\ce{Cl-}/\ce{Br-}/\ce{I-}). The viscoelastic and diffusion properties of the resulting materials are characterized by different rheological and calorimetric analyses, evidencing a strengthening of the gel network upon reinforcing the hard character of the cation and the soft character of the anionic counterion.

@@

The phosphonate terminated polymer \ce{PBr}, which is the common basis to all studied polymer was synthesized according to our previously reported methodology; allowing us to obtain a well-controlled linear polymer ($Mw= \SI{5614}{\dalton}$, $Mw/Mn = 1.08$; S2, S3, Figure S1) .9, 10 All subsequent transformations of the polymer structures were achieved using the same initial batch of material. Nucleophilic addition of N-Methylimidazole or N-methylpyrrolidine to a heated solution of \ce{PBr} in THF afforded the corresponding poly(imidazolium) and poly(pyrrolidinium) compounds. In these cases, as an inherent consequence of the structure of the starting material, charge neutrality is provided by bromide counter ions, and the polymer will be referred to as \ce{PIm+Br-} and \ce{PPyr+Br-}, respectively (Scheme~\ref{sch:synthesis}).

@@

Calorimetric data were then recorded in order to characterize the effect of counterion substitution. In solids, melting points are inherently connected to the reticular enthalpy of the crystal lattice, hence primarily to ion pairing strength.11 This last parameter is directly related to the diffuse and hindered nature of the interacting charges, bulkier and softer ions providing in the most general case weaker interactions.12 For typical ``ionic liquids'' such as imidazolium or pyrollidinium halides, melting points will thus gradually increase as the size of the halide counterion decreases, ie $\ce{I-}<\ce{Br-}<\ce{Cl-}<\ce{F-}$.13 Although the trend is less obvious in amorphous polymeric systems, in which phase transitions result from the interplay of multiple parameters, ionic substitution is however expected to have an impact on the melting temperature of the material.14

@@

Rheological measurements. In order to refine these visual observations, we performed rheological studies of the water-swelling materials. Briefly, data were recorded at \SI{25}{\degree} with an AR 1000 rheometer (TA Instruments) in a cone-plate geometry of radius \SI{40}{\milli\metre}, an angle of \SI{2}{\degree} and a truncation of \SI{58}{\micro\metre}.15, 16 The sample was placed on the plate, the cone was then lowered to the measuring position, spreading the sample in the process. Excess material was removed, so that the sample exactly filled the gap. Due to Weissenberg effect at high shear rate, no preshear was applied before measurement. To minimize water absorption, the geometry was covered with a solvent trap, using light mineral oil as a liquid seal between the rotor and the cap. After one minute of equilibration we performed an oscillatory frequency sweep at small amplitude (strain amplitude $\gamma=0.1\%$) and then an oscillatory strain sweep at fixed frequency ($f=\SI{1}{\hertz}$). For fluoride counterions as well as \ce{PIm+Cl-}, the rheological profile was that of pure water which confirms the visual observation. However as shown in Figure~\ref{fig:strain}, all other samples are solidlike ($G^\prime \gg G^{\prime\prime}$) at small strain and flow at large strains ($G^{\prime\prime} \gg G^\prime$). Solidlike behavior at low strain was confirmed for all accessible frequencies (see supplementary material S6 Fig. 1). We define the critical strain value $\gamma_c$ between these two regimes when $G^{\prime\prime}$ reaches a maximum. We checked that the linear mechanical properties are unchanged by the flow history given a few minutes of rest. %To understand the influence of the ion pairing on the mechanical properties, we compute the cohesive energy (equation 1) that should probe the strength of the physical supramolecular bonds.15,16

\begin{figure}
\includegraphics{strain.pdf}
\caption{Storage modulus $G^\prime$ (filled symbols) and loss modulus $G^{\prime\prime}$ (open symbols) measured through oscillatory shear experiments for (a) \ce{PPyr+Cl-} (\textcolor{lightgray}{$\bullet$}), \ce{PPyr+Br-} (\textcolor{gray}{$\blacksquare$}) and \ce{PPyr+I-} ($\blacktriangle$), (b) \ce{PIm+I-} ($\blacktriangledown$) and \ce{PPyr+I-} ($\blacktriangle$) and (c) \ce{PIm+Br-} (\textcolor{gray}{$\blacklozenge$}) and \ce{PPyr+Br-} (\textcolor{gray}{$\blacksquare$}). $G^\prime$  and $G^{\prime\prime}$ measured during an amplitude sweep at a fixed frequency $f=\SI{1}{\hertz}$ are plotted against the strain amplitude $\gamma$. All samples are at 8\%wt }
\label{fig:strain}
\end{figure}

@@

The above dependence of the gel parameters on the nature of the counterions could be rationalized on the basis of ion-pairing processes. The unsolubility/unswellability of \ce{PIm+F-}, \ce{PPyr+F-} and \ce{PIm+Cl-} suggest that ion pairs are not efficiently dissociated in aqueous medium, thus hindering the solvation process. By comparison, \ce{PIm+Br-}, \ce{PIm+I-}, \ce{PPyr+Br-} and \ce{PPyr+I-} form homogeneous materials when suspended in water, which one can correlate with efficient dissociation of the halide counterions from the polymer main chain is fully consistency with decreased ionic bond strength. Such dissociation, in turns makes it possible for the ``hard'' phosphonate group on each polymer chain's end to interact with the ``free'' cationic groups of the side chains, which we showed in a previous paper to be at the origin of the gel formation mechanism.9

@@

As already mentioned (vide supra) another striking feature of these gels is that in similar conditions poly(pyrrolidinium) based gels are systematically much stronger than those formed with poly(imidazolium). Indeed, although both cationic in nature, pyrrolidinium and imidazolium ions have markedly different properties: whereas pyrrolidinium features a ``hard'' positive charge very localized on the quaternary nitrogen, imidazolium has a much more delocalized ``softer'' charge; consequently, according to the Hard and Soft Acid and Bases (HSAB) theory, stronger association is expected for the latter with the ``soft'' halide counterions.17 As a consequence, for a given halide, dissociation is less favored in the case of poly(imidazolium), as clearly illustrated by the difference in solvation between \ce{PIm+Cl-} and \ce{PPyr+Cl-}. Conversely, association of the lateral cations with the ``hard'' phosphonate moieties at the polymer chain’s end is expected to be stronger with pyrrolidinium, hence a systematically larger elastic modulus. The interplay of these two effects contributes in making poly(pyrrolidinium) based gels much stronger than their poly(imidazolium) counterparts; comparison between \ce{PIm+Br-} and \ce{PPyr+Br-} is particularly illustrative of this trend, as their $G^\prime$ value (13 and \SI{8100}{\pascal}, resp.) differ by almost three orders of magnitude.

@@

>>>>

\section{Introduction}
Since the pioneering work of Wichterle et al.,\cite{Wichterle1960} which established their relevance in a biomedical context, polymer based hydrogels have never ceased to be a very active field of research.\cite{Calo2015,Hoffman2001} They have recently experienced a burst of interest among the biomedical community as controlled drug delivery cargos or scaffolds for wound healing or tissues repair.\cite{Vashist2014,Ratner2004} Polyelectrolytes are being extensively used in this particular context.\cite{Rosso2003,BinImran2014} Their gel formation mechanism often involves reversible-by-nature electrostatic interactions, which can be used for instance to trigger ``smart'' release of bioactive substances.\cite{Khare1993,Lockwood2007} Besides, poly(cationic) gel have been reported to combine scaffold properties for cell’s adherence and growth with antimicrobial activity, and are therefore increasingly used in clinical applications.\cite{Hoffman2001,Landers2002} In this framework, injectability is a key feature, as it provides an easier way to gel delivery in vivo; thus, reversible shear thinning biocompatible hydrogels are a particularly sought after class of materials.\cite{Yu2008}

We have recently reported on a highly reversible poly(cation) based hydrogel,\cite{Srour2014} which formation relies on an new concept of supramolecular  electrostatic interaction. Briefly, the atom transfer radical polymerisation (ATRP) is initiated by a phosphonate terminated ATRP initiator, eventually affording an anion terminated poly(cationic) polymer. The resulting poly(cationic) polymer, when dispersed in an aqueous medium, stabilizes hydrogel formation through the occurrence of a head(anion)-to-body(cations) supramolecular network. This highly dynamic electrostatic network provides the resulting gel with spectacular mechanical and self-healing properties. Moreover the interaction can be disrupted by various chemical stimuli, such as pH or ionic strength.

In the present report, we show that the mechanical/rheological properties of such smart hydrogels can be tuned to an extremely large extent by playing with the nature (hardness/softness) of the cationic repeating unit and the ion-paring strength of the associated counterion (\ce{F-}/\ce{Cl-}/\ce{Br-}/\ce{I-}). The viscoelastic and diffusion properties of the resulting materials are characterized by different rheological and calorimetric analyses, evidencing a strengthening of the gel network upon reinforcing the hard character of the cation and the soft character of the anionic counterion.

@@

The phosphonate terminated polymer \ce{PBr}, which is the common basis to all studied polymer was synthesized according to our previously reported methodology; allowing us to obtain a well-controlled linear polymer ($Mw= \SI{5614}{\dalton}$, $Mw/Mn = 1.08$; S2, S3, Figure S1) .\cite{Srour2014,Appukuttan2012} All subsequent transformations of the polymer structures were achieved using the same initial batch of material. Nucleophilic addition of N-Methylimidazole or N-methylpyrrolidine to a heated solution of \ce{PBr} in THF afforded the corresponding poly(imidazolium) and poly(pyrrolidinium) compounds. In these cases, as an inherent consequence of the structure of the starting material, charge neutrality is provided by bromide counter ions, and the polymer will be referred to as \ce{PIm+Br-} and \ce{PPyr+Br-}, respectively (Scheme~\ref{sch:synthesis}).

@@

Calorimetric data were then recorded in order to characterize the effect of counterion substitution. In solids, melting points are inherently connected to the reticular enthalpy of the crystal lattice, hence primarily to ion pairing strength.\cite{Sherman1932} This last parameter is directly related to the diffuse and hindered nature of the interacting charges, bulkier and softer ions providing in the most general case weaker interactions.\cite{Krossing2006} For typical ``ionic liquids'' such as imidazolium or pyrollidinium halides, melting points will thus gradually increase as the size of the halide counterion decreases, ie $\ce{I-}<\ce{Br-}<\ce{Cl-}<\ce{F-}$.\cite{Dean2010} Although the trend is less obvious in amorphous polymeric systems, in which phase transitions result from the interplay of multiple parameters, ionic substitution is however expected to have an impact on the melting temperature of the material.\cite{Orler1994}

@@

Rheological measurements. In order to refine these visual observations, we performed rheological studies of the water-swelling materials. Briefly, data were recorded at \SI{25}{\degree} with an AR 1000 rheometer (TA Instruments) in a cone-plate geometry of radius \SI{40}{\milli\metre}, an angle of \SI{2}{\degree} and a truncation of \SI{58}{\micro\metre}.\cite{Macosko1994,Larson1999} The sample was placed on the plate, the cone was then lowered to the measuring position, spreading the sample in the process. Excess material was removed, so that the sample exactly filled the gap. Due to Weissenberg effect at high shear rate, no preshear was applied before measurement. To minimize water absorption, the geometry was covered with a solvent trap, using light mineral oil as a liquid seal between the rotor and the cap. After one minute of equilibration we performed an oscillatory frequency sweep at small amplitude (strain amplitude $\gamma=0.1\%$) and then an oscillatory strain sweep at fixed frequency ($f=\SI{1}{\hertz}$). For fluoride counterions as well as \ce{PIm+Cl-}, the rheological profile was that of pure water which confirms the visual observation. However as shown in Figure~\ref{fig:strain}, all other samples are solidlike ($G^\prime \gg G^{\prime\prime}$) at small strain and flow at large strains ($G^{\prime\prime} \gg G^\prime$). Solidlike behavior at low strain was confirmed for all accessible frequencies (see supplementary material S6 Fig. 1). We define the critical strain value $\gamma_c$ between these two regimes when $G^{\prime\prime}$ reaches a maximum. We checked that the linear mechanical properties are unchanged by the flow history given a few minutes of rest. %To understand the influence of the ion pairing on the mechanical properties, we compute the cohesive energy (equation 1) that should probe the strength of the physical supramolecular bonds.15,16

\begin{figure}
\includegraphics{strain.pdf}
\caption{Storage modulus $G^\prime$ (filled symbols) and loss modulus $G^{\prime\prime}$ (open symbols) measured through oscillatory shear experiments plotted against the strain amplitude $\gamma$ for (a) \ce{PPyr+Cl-} (\textcolor{lightgray}{$\bullet$}), \ce{PPyr+Br-} (\textcolor{gray}{$\blacksquare$}) and \ce{PPyr+I-} ($\blacktriangle$), (b) \ce{PIm+I-} ($\blacktriangledown$) and \ce{PPyr+I-} ($\blacktriangle$) and (c) \ce{PIm+Br-} (\textcolor{gray}{$\blacklozenge$}) and \ce{PPyr+Br-} (\textcolor{gray}{$\blacksquare$}). The fixed frequency is $f=\SI{1}{\hertz}$. All samples are at 8\%wt }
\label{fig:strain}
\end{figure}

@@

The above dependence of the gel parameters on the nature of the counterions could be rationalized on the basis of ion-pairing processes. The unsolubility/unswellability of \ce{PIm+F-}, \ce{PPyr+F-} and \ce{PIm+Cl-} suggest that ion pairs are not efficiently dissociated in aqueous medium, thus hindering the solvation process. By comparison, \ce{PIm+Br-}, \ce{PIm+I-}, \ce{PPyr+Br-} and \ce{PPyr+I-} form homogeneous materials when suspended in water, which one can correlate with efficient dissociation of the halide counterions from the polymer main chain is fully consistency with decreased ionic bond strength. Such dissociation, in turns makes it possible for the ``hard'' phosphonate group on each polymer chain's end to interact with the ``free'' cationic groups of the side chains, which we showed in a previous paper to be at the origin of the gel formation mechanism.\cite{Srour2014}

@@

As already mentioned (vide supra) another striking feature of these gels is that in similar conditions poly(pyrrolidinium) based gels are systematically much stronger than those formed with poly(imidazolium). Indeed, although both cationic in nature, pyrrolidinium and imidazolium ions have markedly different properties: whereas pyrrolidinium features a ``hard'' positive charge very localized on the quaternary nitrogen, imidazolium has a much more delocalized ``softer'' charge; consequently, according to the Hard and Soft Acid and Bases (HSAB) theory, stronger association is expected for the latter with the ``soft'' halide counterions.\cite{Goossens2009} As a consequence, for a given halide, dissociation is less favored in the case of poly(imidazolium), as clearly illustrated by the difference in solvation between \ce{PIm+Cl-} and \ce{PPyr+Cl-}. Conversely, association of the lateral cations with the ``hard'' phosphonate moieties at the polymer chain’s end is expected to be stronger with pyrrolidinium, hence a systematically larger elastic modulus. The interplay of these two effects contributes in making poly(pyrrolidinium) based gels much stronger than their poly(imidazolium) counterparts; comparison between \ce{PIm+Br-} and \ce{PPyr+Br-} is particularly illustrative of this trend, as their $G^\prime$ value (13 and \SI{8100}{\pascal}, resp.) differ by almost three orders of magnitude.

@@

\bibliography{PALSE.bib}
